Gametic Selection, Meiotic Drive, Sex Ratio Bias, and Transitions Between Sex Determination Systems

### 2017-02-27 LaTeX/Turnover.tex
started looking at weak recombination but need sleep

\begin{threeparttable}[ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{\female} w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + 2(1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^\male (1 - R)\}/ \{2q \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male\} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{\female}) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + 2p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha^\male) (1 - R)\}/ \{ 2q \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^{\male}/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha^{\male})/ \{ \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+2(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}\alpha^\female(1-R)\}/ \{ 2(1-q) \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+2\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha^\female)(1-R)\}/ \{2(1-q) \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} \alpha^{\female}/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha^{\female})/ \{\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
      \item $R$ is the probability of recombination between loci \textbf{A} and \textbf{M}.
      \item See Table \ref{tab:meanfitnesses} for expressions of mean fitnesses.
%      \item $\bar{w}_H^\female = p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ is the mean fitness of female gametes during gametic competition
%      \item $\bar{w}_H^\male = \bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ is the mean fitness of male gametes during gametic competition
%       \item $\bar{w}^\female = \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}$ is the mean fitness of females during diploid competition
%       \item $\bar{w}^\male = \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\}$ is the mean fitness of males during diploid competition
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}
\\

@@

\begin{table}[ht]
\centering
\smallskip
\caption{mean fitnesses in resident ($M$ fixed, XY sex determination) }
\begin{tabular}{l l }
\hline\hline
  Sex \& Life Cycle Stage & Mean Fitness \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  female gametes ($\bar{w}_H^\female$) & 
  $p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  male gametes ($\bar{w}_H^\male$) & 
  $\bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  females ($\bar{w}^\female$) & 
  $\begin{array}{l}  \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + \\
  (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + \\
  p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + \\
  (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}
  \end{array} 
  $ \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  males ($\bar{w}^\male$) & 
  $\begin{array}{l} \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + \\
  (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + \\
  p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + \\
  (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\} 
  \end{array}
  $ \\ [0.5ex]  \noalign{\vskip 0.5ex}
  \hline \hline
  \label{tab:meanfitnesses}
 \end{tabular}
\end{table}

@@

Here, we determine whether a rare neo-Y or neo-W allele spreads when rare, which occurs when $\lambda > 1$. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$, see table \ref{tab:haplotype_growth} for $\lambda_{mi}$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ 
\frac{p^\female_X w^\female_A w^\male_a (1-\alpha^\male)}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{mA}-1)} + 
%p^\female_X \frac{w^\female_A}{\bar{w}^\female_H} \frac{w^\male_a(1-\alpha^\male)}{\bar{w}^\male_H} \frac{1}{\lambda_{mA}-1} + 
\frac{(1 - p^\female_X) w^\female_a w^\male_A \alpha^\male}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\male_{Aa}}{\bar{w}^\male} < 1,
\end{equation}

@@

\noindent
\begin{equation}\label{eq:neoWR}
R \left[ 
\frac{\bar{p}^\male w^\male_A w^\female_a (1-\alpha^\female)}{ \bar{w}^\male_H \bar{w}^\female_H (\lambda_{mA}-1)} + 
\frac{(1 - \bar{p}^\male) w^\male_a w^\female_A \alpha^\female}{ \bar{w}^\male_H \bar{w}^\female_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\female_{Aa}}{\bar{w}^\female} < 1,
\end{equation}

>>>>

\begin{threeparttable}[ht]
\centering
\smallskip
%\caption{Parameters determining invasion (equation \ref{eq:charpoly_neoY}) for neo-Y or neo-W alleles}
\caption{Parameters determining invasion of mutant neo-Y and neo-W alleles into an ancestrally XY system}
\begin{tabular}{l}
\hline\hline
   \noalign{\vskip 0.5ex}
   neo-Y ($k=0$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{p_{X}^{\female} w_{A}^{\female} w_{A}^{\male} w_{AA}^{\male} + 2(1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^\male (1 - R)\}/ \{2q \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male\} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-p_{X}^{\female}) w_{a}^{\female} w_{a}^{\male} w_{aa}^{\male} + 2p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male}(1 - \alpha^\male) (1 - R)\}/ \{ 2q \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-p_{X}^{\female}) w_{a}^{\female} w_{A}^{\male} w_{Aa}^{\male} \alpha^{\male}/ \{ q \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R p_{X}^{\female} w_{A}^{\female} w_{a}^{\male} w_{Aa}^{\male} (1 - \alpha^{\male})/ \{ q \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\male \}$\\ [1ex] \hline 
  \noalign{\vskip 0.5ex}
  neo-W ($k=1$) \\ [0.5ex] \hline \noalign{\vskip 1ex}
  $\lambda_{mA}=\{\bar{p}^{\male} w_{A}^{\male} w_{A}^{\female} w_{AA}^{\female}+2(1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female}\alpha^\female(1-R)\}/ \{ 2(1-q) \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\lambda_{ma}=\{(1-\bar{p}^{\male}) w_{a}^{\male} w_{a}^{\female} w_{aa}^{\female}+2\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female}(1-\alpha^\female)(1-R)\}/ \{2(1-q) \bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$ \\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{mA}=R (1-\bar{p}^{\male}) w_{a}^{\male} w_{A}^{\female} w_{Aa}^{\female} \alpha^{\female}/ \{(1-q)\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \} $\\ [0.5ex] \noalign{\vskip 0.5ex}
  $\rho_{ma}=R\bar{p}^{\male} w_{A}^{\male} w_{a}^{\female} w_{Aa}^{\female} (1-\alpha^{\female})/ \{(1-q)\bar{w}_{H}^{\female} \bar{w}_{H}^{\male} \bar{w}^\female \}$\\ [1ex] 
  \hline \hline 
   \end{tabular}
      \begin{tablenotes}
      \scriptsize
      \item $\bar{p}^{\male}=p_{Y}^{\male}q+p_{X}^{\male}(1-q)$ is the average frequency of the $A$ allele among X- and Y-bearing male gametes.
      \item $R$ is the probability of recombination between loci \textbf{A} and \textbf{M}.
      \item See Table \ref{tab:meanfitnesses} for expressions of mean fitnesses.
%      \item $\bar{w}_H^\female = p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ is the mean fitness of female gametes during gametic competition
%      \item $\bar{w}_H^\male = \bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ is the mean fitness of male gametes during gametic competition
%       \item $\bar{w}^\female = \{ p_X^\female w_A^\female (1-q) p_X^\male w_A^\male w_{AA}^\female + (1 - p_X^\female) w_a^\female (1-q) p_X^\male w_A^\male w_{Aa}^\female + p_X^\female w_A^\female (1-q) (1 - p_X^\male) w_a^\male w_{Aa}^\female + (1-p_X^\female) w_a^\female (1-q) (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}$ is the mean fitness of females during diploid competition
%       \item $\bar{w}^\male = \{ p_X^\female w_A^\female q p_Y^\male w_A^\male w_{AA}^\male + (1 - p_X^\female) w_a^\female q p_Y^\male w_A^\male w_{Aa}^\male + p_X^\female w_A^\female q (1 - p_Y^\male) w_a^\male w_{Aa}^\male + (1-p_X^\female) w_a^\female q (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\}$ is the mean fitness of males during diploid competition
    \end{tablenotes}
  \label{tab:haplotype_growth}
\end{threeparttable}
\\

@@

\begin{table}[ht]
\centering
\smallskip
\caption{mean fitnesses in resident ($M$ fixed, XY sex determination) }
\begin{tabular}{l l }
\hline\hline
  Sex \& Life Cycle Stage & Mean Fitness \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  female gametes ($\bar{w}_H^\female$) & 
  $p_X^\female w_A^\female + (1-p_X^\female) w_a^\female$ \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  male gametes ($\bar{w}_H^\male$) & 
  $\bar{p}^{\male} w_A^\male + (1-\bar{p}^{\male}) w_a^\male$ \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  females ($\bar{w}^\female$) & 
  $\begin{array}{l}  \{ p_X^\female w_A^\female p_X^\male w_A^\male w_{AA}^\female + \\
  (1 - p_X^\female) w_a^\female p_X^\male w_A^\male w_{Aa}^\female + \\
  p_X^\female w_A^\female (1 - p_X^\male) w_a^\male w_{Aa}^\female + \\
  (1-p_X^\female) w_a^\female (1 - p_X^\male) w_a^\male w_{aa}^\female \} / \{ \bar{w}_H^\female \bar{w}_H^\male\}
  \end{array} 
  $ \\ [0.5ex] \hline  \noalign{\vskip 0.5ex}
  males ($\bar{w}^\male$) & 
  $\begin{array}{l} \{ p_X^\female w_A^\female  p_Y^\male w_A^\male w_{AA}^\male + \\
  (1 - p_X^\female) w_a^\female  p_Y^\male w_A^\male w_{Aa}^\male + \\
  p_X^\female w_A^\female  (1 - p_Y^\male) w_a^\male w_{Aa}^\male + \\
  (1-p_X^\female) w_a^\female  (1 - p_Y^\male) w_a^\male w_{aa}^\male \} / \{ \bar{w}_H^\male \bar{w}_H^\male\} 
  \end{array}
  $ \\ [0.5ex]  \noalign{\vskip 0.5ex}
  \hline \hline
  \label{tab:meanfitnesses}
 \end{tabular}
\end{table}

@@

Here, we determine whether a rare neo-Y or neo-W allele spreads when rare, which occurs when $\lambda > 1$. 
If the average change in frequency of the two haplotypes that carry the $m$ allele ($Am$ and $am$) is positive, invasion will always occur (i.e., if $\left\{(\lambda_{mA}-1)+ (\lambda_{ma}-1) \right\}/2 > 0$ then $\lambda > 1$, see table \ref{tab:haplotype_growth} for $\lambda_{mi}$). 
If neither haplotype increases in frequency ($\lambda_{mA}, \lambda_{ma} < 1$), the $m$ allele will not invade. 
Otherwise, the new sex-determining allele increases in frequency on one \textbf{A} background and declines on the other, and invasion requires
\begin{equation}\label{eq:neoYR}
R \left[ 
\frac{p^\female_X w^\female_A w^\male_a (1-\alpha^\male)}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{mA}-1)} + 
%p^\female_X \frac{w^\female_A}{\bar{w}^\female_H} \frac{w^\male_a(1-\alpha^\male)}{\bar{w}^\male_H} \frac{1}{\lambda_{mA}-1} + 
\frac{(1 - p^\female_X) w^\female_a w^\male_A \alpha^\male}{ \bar{w}^\female_H \bar{w}^\male_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\male_{Aa}}{q\bar{w}^\male} < 1,
\end{equation}

@@

\noindent
\begin{equation}\label{eq:neoWR}
R \left[ 
\frac{\bar{p}^\male w^\male_A w^\female_a (1-\alpha^\female)}{ \bar{w}^\male_H \bar{w}^\female_H (\lambda_{mA}-1)} + 
\frac{(1 - \bar{p}^\male) w^\male_a w^\female_A \alpha^\female}{ \bar{w}^\male_H \bar{w}^\female_H (\lambda_{ma}-1)} 
\right] 
\frac{w^\female_{Aa}}{(1-q)\bar{w}^\female} < 1,
\end{equation}
